# Introduction to data types 

We will start with some of the most basic data units in R and work our way up to some of the most complex. The most basic unit we will deal with are *data types*. There are many data types in R, but we will mostly deal with four: **characters**, **numerics**, **integers**, and **logicals**.

## Character

A **character** data type is used to store text.

```{r}
'Hello, World'
```

If you run this code, you will have just created a character data type.

We can use the `class` function to get the data type or data structure. More
about functions later! For now, you should know that you can ask R about the
data type by enclosing it in parentheses after the word `class`.

```{r, eval = T}
class('Hello, World')
```

You will notice R tells you the class is **"character"**.

## Numeric

A **numeric** data type is used to store real numbers.

```{r}
4.78
class(4.78)
```

An **integer** is a special subset of the numeric data type used to store integers. When we want to tell R we want a number as an integer, we type an "L" after the number.

```{r}
4L
class(4L)
```

## Logical

A **logical** type takes two possible values: "TRUE" or "FALSE". Note these values are always capitalized. R also recognizes shortforms for logicals "T" and "F".

```{r}
TRUE
class(TRUE)
class(F)
```

## Checking the data type

We can also ask R whether a data type is of a specific type using the **"is"** set of functions. We use this function by typing is.DATATYPE() and enclosing the data type we aren't sure about in the parentheses. For example `is.numeric(3L)` asks R whether 3L is numeric and returns a logical datatype specifying whether the statement is TRUE (yes, it's a numeric) or FALSE (no, it's not a numeric). Here are some examples. Before running the code, try to predict what classes R will return. You should also try some others yourself!

```{r}
is.character(TRUE)
is.logical(TRUE)
is.character('TRUE')
```
