---
title: "Take-home Challenge 1"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 80), tidy=TRUE)
```

### Take-home Challenge 1

This challenge is due **Thursday, Jan. 29th at 11:59 pm**. It will be graded out of 10 points (the marks distribution is outlined in each section). Your grade on the assignment will make up approximately 3--5% of your final grade, which will depend on how many additional Take-home Challenges are assigned this semester and how much work you are required to do in each. This will likely be the challenge with the lightest workload.

You can either complete your assignment using this .Rmd document or using the TakeHomeChallengeFormatting.Rmd. If you use this document, please ensure you include the necessary headings and copy and replace the YAML in this document with the YAML from TakeHomeChallengeFormatting.Rmd (everything at the head of the  document between "---" and "---") and change the title and author, e.g.:

---
title: "Take Home Challenge X"
author: "YOUR NAME HERE"
---

Required submissions for this Take Home Challenge: 

* your .Rmd
* your output knitted as an HTML and saved to a PDF

### Your task

You are a summer RA reviewing some data the lab collected the previous summer. The dataset contains the physiological performance of five animal species subjected to an experiment on physiological performance. The authors collected metabolic, temperature, and movement speed data from separate trials. The columns are:

`individual_id`: unique identification number of the individual animal
`species`: species name and genus to which the individual belongs
`temperature_C`: the body temperature of the individual at the time of the trial in Celsius
`metabolic_rate_ml02_g_hr`: oxygen consumed by the individual during the trial in milliliters per gram of body mass per hour
`spring_speed_cm_s`: movement speed of the individual during the trial

Your supervisor has asked you some very specific questions about the data and you decided to give them a reproducible report with the answers.

1. **(2 pts)** Create a new folder called `BIOL3401_THC1` somewhere on your computer and save or copy the `THC1_data.csv` file to the folder (note: you should do this step *outside* of R). Write the code needed to:

  a. Set the new folder (`BIOL3401_THC1`) as your working directory.
  b. Load the `THC1_data.csv` dataset directly from the folder, assigning it the name `trial_data`.

```{r}

```

2. **(1.5 pts)** Create a new folder called `data` inside the `BIOL3401_THC1` folder (note: you should do this step outside of R). Save (write) the same `trial_data` to the new folder, saving it as `updated_trial_data.csv`. The datasets are identical for now, but you have set up a structure that can be used to automatically save any future updates to the file.

```{r}

```

3. **(1 pt)** List the data *types* of the five columns:

* `individual_id`: ______________
* `species`: ______________
* `temperature_C`: ______________
* `metabolic_rate_ml02_g_hr`: ______________
* `sprint_speed_cm_s`: ______________

4. **(1.5 pts)** Print the first 11 rows of the data using the `head()` function.

```{r}

```

5. **(1.5 pts)** Create a new data frame called `mouse_data` with only observations from the species *Mus musculus*.

```{r}

```

6. **(1.5 pts)** Create a new data frame called `terrestrial_data` that does not include the species *Oncorhynchus mykiss*.

```{r}

```

7. **(1 pt)** Use code to find the following two pieces of data:

  a. The metabolic rate of *Parus major* individual B08.
  b. All data from *Lithobates pipiens* individual F02.

```{r}

```


