---
title: "Data Visualization"
author: "BIOL 3401 Mount Royal University"
date: "Topic 4, Winter 2026"
output: 
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 80), tidy=TRUE)
```


## Data Visualization

One of the major advantages of using R for data analyses is that it is very easy to produce figures that are high quality for publication. Please visit <https://www.r-graph-gallery.com/> and the `ggplot2` cheat sheet on the Google Drive to get an idea for what the options and possibilities are.

`ggplot2` (referred to simply as 'ggplot') is a `tidyverse` package that is basically the gold standard for plotting data when using R. The basic anatomy of a ggplot consists of the following elements:

| Syntax     | Description                                          |
|------------|------------------------------------------------------|
| `ggplot()` | Creates a ggplot                                     |
| `aes()`    | Constructs aesthetic mappings                        |
| `+`        | Adds a component to your plot                        |
| `geom_`    | Geometric layer (e.g., points, lines...)             |

With these basic elements, you can create a high quality figure, and there are many aesthetic components you can add/change as well. The workflow for making a ggplot is typically an iterative process where you gradually refine your work towards your finished project.

# 1 - Set up your R workspace for the day

## Step 1a: Set your working directory

Make sure we're setting our working directory!

```{r, eval = F}
# Set your working directory, then remove "eval = F"!
setwd()
```

## Step 1b: Load packages

```{r, message = F}
# Load tidyverse (ggplot2 is a part of the tidyverse, so you already have it installed)
library(tidyverse) 
```

## Step 1c: Load data

We are using four datasets today. All four are .csv files. We are going to use them to create four different plots, visualizing the patterns and adding new ggplot2 elements as we go. The datasets are as follows:

* **cell_metabolism.csv**: Contains 150 observations from three human cell types. We are going to use the data to visualize the relationship between cell size and metabolic rate.
  * cell_type: The type of cell (epithelial, neuron, or muscle)
  * length_um: The length of the cell in micrometers
  * metabolic_rate_pmol_min: The metabolic rate of the cell in picomoles per minute
* **cortisol_stress.csv**: Contains 240 observations from an experiment where human subjects experienced stressor "treatments", and then blood samples were taken to measure their circulating blood cortisol (stress hormone). We are going to use the data to visualize the relationship between human exposure to stressors and stress levels.
  * treatment: The stressor the subject was exposed to (Control, Heat stress, or Diet restriction)
  * sex: M (male) or F (female) subjects
  * cortisol_ng_ml: The concentration of circulating blood cortisol after exposure to the stressor in nanograms per millileter
* **exercise_recovery.csv**: Contains 63 observations of human heart rates following exercise. We are going to use the data to visualize cardiovascular recovery over time.
  * time_min: Time since stopping exercise in minutes
  * heart_rate_bpm: Heart rate of the subject in beats per minute
  * fitness_level: Relative fitness level of the subject (Low, Moderate, or High)
* **fish_stocking.csv**: Contains 8 observations of fish survival under different hatchery stocking densities (fish per m^2). We are going to use the data to compare survival by stocking density.
  * density: The density treatment where the fish were stocked (Low, Medium, High, or Very High)
  * environment: The setting where the fish was stocked (Artificial tank or Natural pond)
  * n_stocked: Number of fish stocked in the treatment
  * n_survived: The number of fish that survived out of those that were stocked
  * survival_proportion: The proportion of surviving fish out of those that were stocked
  
Let's load the datasets now! Finish the `read_csv()` functions.

```{r}
# Load the cell_metabolism.csv data
cell_data <- read_csv()
# Load the cortisol_stress.csv data
cortisol_data <- read_csv()
# Load the exercise_recovery.csv data
recovery_data <- read_csv()
# Load the fish_stocking.csv data
fish_data <- read_csv()
```

# 2 - Construct a basic scatterplot

We are going to start with cell_data. We are going to construct a scatterplot to visualize the relationship between cell size and metabolic rate.

## Step 2a: ggplot2 basics

The first layer to specify when making figures is the global axes. These axes will apply to all layers that follow this first one. This is typically where you want to specify your main axes (in the aesthecic) and the dataset.

```{r}
ggplot(cell_data, aes(x = length_um, y = metabolic_rate_pmol_min))
```

As we can see, there are axes, but no data has been plotted! This is because we haven't told R in what *layers* (`geoms`) we want to plot it in. All we have specified is what to put on the x and y axes.

We are interested in the relationship between cell size and metabolic rate, both of which are continuous variables. This is a perfect application for a `geom_point()` (scatterplot).

```{r}
ggplot(cell_data, aes(x = length_um, y = metabolic_rate_pmol_min)) + 
  geom_point()
```

Something doesn't seem quite right here. Why do we have a stack of points lined up around zero length?

This is actually quite a common problem when we plot data like sizes, masses, distances, volumes, etc. because they can often vary by orders of magnitude. The issue is that the length_um variable needs to be *rescaled* to prevent the very large cells from overwhelming the plot. We are going to adjust the axis to a *natural log scale* by using the `log()` function. We could use `log()` within the `mutate()` function to add a column to the data with logged lengths, but ggplot2 actually makes it possible to do that within the aesthetic.

```{r}
ggplot(cell_data, aes(x = log(length_um), y = metabolic_rate_pmol_min)) + 
  geom_point()
```

That's much better! We can now see the small cells (~ 100 micrometers) as well as the larger cells (~ 8 millimeters).

Let's also add a "trend line" to the plot. The simplest trend line we can add is a linear trend line. We can tell the relationship here is not exactly linear 

- Rename the axes to specify logs, etc.
- Colour the points by group

```{r}
snailed_it <- snail_survival %>%
  mutate(temp = as_factor(temp))

ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_boxplot()
```

That's better!

#### Layering Geoms

We have our first box plot! It's a good idea to add data distributions so that we can see everything, as some people feel like box plots don't sufficiently display data. We can stack another geometry argument on top of our box plot to display all our data points; in this case `geom_point()`.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_boxplot() + 
  geom_point()
```

Well... there are points. But it looks terrible. We can make the points "jitter" so they aren't placed all on top of each other.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_boxplot() + 
  geom_jitter()
```

Better. But this makes it really hard to tell which dots are which, because they're all black!

`fill` changes the colour of the inside of an object, while `colour` changes the outline. Let's add colour to our aesthetics. To do this, we specify which column in our data frame will be used to specify the different colour groups.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp, colour = temp)) +
  geom_boxplot() + 
  geom_jitter()
```

Terrible. Do you notice how the lines on the box plot changed too?? 

Plus, we can't see data because everything is the same colour! Luckily, we can change the aesthetics within individual geoms - so let's move the `colour` argument to the `geom_jitter` function.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_boxplot() + 
  geom_jitter(aes(colour = temp))
```

#### Organizing layers

Getting there! But, it's annoying that the points are above the box plot and it looks messy. We can adjust that by changing the order of the layers while building the plot. The order of our geom arguments is the order they will be layered on our plots.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot() 
```

#### Alpha

Now the box plot is back in front of the points, but it's still blocking our data. To fix this, we can change the opacity with an `alpha` argument in our box plot geometry. An alpha of 1 is completely opaque whereas an alpha of 0 is transparent. Let's just step back a little on the opacity of the box plot:

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) 
```

#### Changing colours

Looks good now! But what if we want to change our colours? You can make a colour variable that you associate with your data, which is a fun way to practice creating variables again. Since temperature usually has pretty common colours associated with it (blue = cold, red = hot), we can make the colours of our boxplots match our treatment!

Colour is an effective tool in visual story telling, but it's also very subjective and we should always be mindful of how people interpret colour. There are some great colour packages out there that take into account how different palettes are interpreted by people with differing abilities to distinguish colours (try the packages `viridis`, `colorBlindness`, or `RColorBrewer`).

```{r}
my_cols <- c("10" = "lightskyblue", "15" = "goldenrod", "20" = "tomato")
```

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols)
```


### **Mini Challenge 1**

1. Pick any other two hex colour code (websites like https://htmlcolorcodes.com/ can help you pick your codes, or Googling 'R colour palettes') and create a new colour object called `my_cols_2`. Use these colours on the previous plot.


Use the code block below to code your answer! 
```{r}

```






###-----------------------------------------------------

```{r}
my_cols_2 <- c("10" = "#2B827B", "15" = "#827B2B", "20" = "#822B32")

ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols_2) +
  scale_fill_manual(values = my_cols_2)
```



#### Themes

One of the easiest ways of making your figures look more professional/less "cookie cutter" is to change the theme. I like a cleaner, simpler look. There are a few different theme options that you can explore, but I prefer to start with `classic()`.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  theme_classic()
```

We can also change where the legend is positioned. Currently it is off to the right side but we can change that with the `theme()` argument.

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  theme_classic() + 
  theme(legend.position = "top")
```

#### Labelling the axes

While the column names in our data set are easy to code with, they don't look great on the axes, do they? We can re-label the axes so they are easier and clearer to read.

### **Mini Challenge 2**

1. Change the code below to so the placeholder labels are meaningful a.k.a. communicate the correct information for each label.

2. Also change the code below to get rid of the legend, rather than having it at the top of the figure (hint: how can you *ask for help in R*?)

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  theme_classic() +
  theme(legend.position = "top") + 
  labs(x = "X axis label here", y = "Y axis label here", 
       color = "Colour label here", 
       fill = "Fill label here") 
```










###-----------------------------------------------------


```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  theme_classic() +
  theme(legend.position = "none") + 
  labs(x = "Temperature (°C)", y = "Percent Survival (%)")
```

#### Faceting

This is an element of visually subsetting your data, again, without messing up your raw data. We can separate the way our data is displayed using a facet. A `facet_wrap` is used for separating by one variable, and a `facet_grid` is used if you want to separate by 2 variables.

We could want to see what the interactive effect of temperature and humidity is on survival:

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  facet_wrap(~ hum) +
  theme_classic() +
  theme(legend.position = "none") + 
  labs(x = "Temperature (°C)", y = "Percent Survival (%)") 
```

That is a NICE figure! 

We are so close to the final touch on this figure. Last thing - there are these weird grey dots on the boxplots. 

These are outliers identified by the `geom_boxplot()` function, but those data points are part of the raw data plotted with `geom_jitter()`. By having BOTH of these data points on the figure, we're double-plotting the same individual(s) when we don't need to. Let's make one small tweak to our boxplot to get rid of this: 

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  facet_wrap(~ hum) +
  theme_classic() +
  theme(legend.position = "none") + 
  labs(x = "Temperature (°C)", y = "Percent Survival (%)") 
```

This new addition to the `geom_boxplot()` function assigned our outliers as NAs so they wouldn't be plotted - but it didn't impact how the `geom_jitter()` function plotted the data! Now we have a nice, clean figure that we've put a lot of effort into.

#### Saving your plot

You can save your plot but first we need to make a plot object, or we can't tell R *which* plot to save very easily. You can assign a plot to a new object with the classic `<-` operator and then to see the plot output you need to call the plot object

```{r}
pretty_boxplot <- ggplot(snailed_it, aes(x = temp, y = perc.surv, fill = temp)) + 
  geom_jitter(aes(colour = temp)) + 
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  scale_color_manual(values = my_cols) +
  scale_fill_manual(values = my_cols) +
  facet_wrap(~ hum) +
  theme_classic() +
  theme(legend.position = "none") + 
  labs(x = "Temperature (°C)", y = "Percent Survival (%)") 

pretty_boxplot
```


There are many options within `ggsave` as to how to save your plot. I recommend saving files in a vector format (.png for example) as these are easier to resize and touch up for different aesthetics in something like Adobe Illustrator. You can also use a .png which be used for changing resolutions. Within `ggsave` you can change the width and height as well to make sure the size and shape of your plot is exactly what you want.

```{r}
ggsave(plot = pretty_boxplot, filename = "pretty_boxplot.pdf")
```

You can also save it with the "Export" button in the Plots pane!


### **More types of visualizations**

#### Histograms

Now that we know how the different layers work within `ggplot`, altering our code to make a histogram is very simple. Instead of `geom_boxplot()` we will use `geom_histogram()`. With a histogram, we also only need to specify the x-axis, as the y-axis will automatically become the count.

Lets make a histogram of the snail percent survival.

```{r}
ggplot(snailed_it, aes(x = perc.surv)) +
  geom_histogram()
```

With histograms we can also change the number of bars with the argument `bins`. We can also change the colour of all our bars using a fill argument, but if we just want to change all the bins and not base the fill on a column in our data frame, this will just go in the `geom_histogram` layer and not in an aesthetic argument (like we did in the box plot).

```{r}
ggplot(snailed_it, aes(x = perc.surv)) +
  geom_histogram(color = "black", fill = "skyblue", bins = 15)
```

### **Mini Challenge 3**

Now lets make this plot a bit prettier. Copy and paste the code above into the code chunk below, and do the following:

1.  Customized x and y labels

2.  *Fill* all the bars with the colour "#127d76".

3.  Add our favourite theme, `theme_classic()`

Use the code chunk below to code your answer!

```{r}

```










#-------------------------------------------------


```{r}
ggplot(snailed_it, aes(x = perc.surv)) +
  geom_histogram(color = "black", fill = "#127d76", bins = 15) + 
  labs(x = "Percent Survival (%)", y = "Count") + 
  theme_classic()
```



#### Interval Plot

Let's make an interval plot of these data to display the means +/- standard error. R can do the mean and standard error calculations for us(!) directly in the plot, so we don't have to bother ourselves with doing extra calculations. 

```{r}
ggplot(snailed_it, aes(x = temp, y = perc.surv)) + 
  stat_summary(fun = mean, geom = "point") + 
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3) + 
  labs(x = "Temperature (°C)", y = "Mean Survival (%)") + 
  theme_bw()
```


#### Scatter Plot

One final plot type you are definitely going to encounter (and probably want to make at some point!) is a scatter plot. This is a great visualization for two continuous variables, so let's use our `iris` data set to try this out. 

```{r}
iris <- read_delim("iris.csv")

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) + 
  geom_point() +
  theme_classic()
```

One final thing you may want to do is add a line of best fit to your plot.

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) + 
  geom_point() +
  theme_classic() +
  geom_smooth(method = "lm")
```

So pretty! Enjoy making beautiful plots.

