---
title: "Introduction to Models"
author: "BIOL 3401 Mount Royal University"
date: "Topic 6, Winter 2026"
output: 
  pdf_document:
    toc: yes
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 80), tidy=TRUE)
```


## Introduction to modelling in R

The most basic form of modelling is a linear model. You will also hear this referred to as a linear regression. We've already discussed how some of the tests you are already familiar with (e.g., ANOVA) are just a different form of linear model. The only real difference is that:

* ANOVA is a *test of difference*. We want to know whether there is a signficant difference between two groups in terms of a discrete variable, **X**, across a response variable, **Y**.
* Linear regression is a *test of relationship*. We want to know whether there is a significant relationship between **X** and **Y**. 

# 1 - Set up your R workspace for the day

## Step 1a: Set your working directory

Make sure we're setting our working directory!

```{r, eval = F}
# Set your working directory, then remove "eval = F"!
setwd()
```

## Step 1b: Install and load packages

We are going to need to load tidyverse. We are also going to install a new package called `car`, which stands for "Companion to Applied Regression".

Using the code block below, install the `car` package, then load `car` and `tidyverse`.

## Step 1c: Load data

Our dataset for this topic (**lizard_data.csv**) comes from a study on lizard running speeds in different habitats. Lizards are ectotherms, which means their physiological performance depends on the external temperature. We will be testing a number of hypotheses relating body size, temperature, age, sex, habitat, and running speed.

Load the dataset by finishing the `read_csv()` function.

```{r, echo = F, message=FALSE}
lizard_data <- read_csv()
```

## Step 2: Test four different hypotheses

Tests of difference:

* body size ~ sex (meets assumptions)
* temperature ~ habitat (violates assumptions)

Plot boxplots and assess assumptions visually, then test with Shapiro and Levene's

Tests of relationship:

* body size ~ age (meets assumptions)
* speed ~ temperature (violates assumptions)

Fit the models, then look at the diagnostic plots (residual plot and qqplots)

## Step 3: Multiple regression

Fit models for:

* body size ~ age
* body size ~ sex
* body size ~ age + sex

Compare r-squareds and assess using AIC.

## Step 4: Interactions

Add interaction between sex and age: body size ~ age*sex

## Step 5: Outliers

Assess outliers using Cook's distance




