---
title: "Generalized Linear Models and Distributions"
author: "BIOL 3401 Mount Royal University"
date: "Topic 7, Winter 2026"
output: 
  pdf_document:
    toc: yes
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 80), tidy=TRUE)
```

# **Learning Objectives**

Today, we will start learning about statistical models. Linear models are one of the simplest and most versatile classes of statistical model. We will add more complexity to our analyses in the weeks to come.

After completing this topic, you should feel comfortable:

1. Comparing and contrasting Generalized Linear Models (GLMs) and simple linear models.
2. Identifying appropriate link functions and distributions (e.g., binomial, Poisson) for different types of  response variables.
3. Fitting GLMs and interpreting results based on the link function.
4. Evaluating GLM model fit.

## Introduction to modelling in R

In many ways, GLMs are more straightforward and more versatile than simple linear models because their assumptions are not quite as restrictive. Instead of checking whether the residuals are normally distributed and independent with homogeneous variance, and ensuring the Y ~ X relationship is linear, we can often just use our understanding of how the data were generated to make the correct modelling choices.

## Set up your R workspace for the day

### Set your working directory

Make sure we're setting our working directory!

```{r}
setwd()
```

### Install and load packages

We are going to need to load `tidyverse` and `car`. We are also going to install two new packages called `DHARMa` and `fitdistrplus`.

Using the code block below, install the two new packages and load all four packages.

```{r}

```

### Load data

We have two datasets for this topic.

* **ivf_data.csv**: Comes from a retrospective study conducted at a fertility clinic. The researchers evaluated predictors of clinical pregnancy (age and hormone treatment) following a single in-vitro fertilization (IVF) cycle. Each row represents a unique patient.
  * `pregnant`: Whether the patient achieved a clinical pregnancy (1 = yes, 0 = no)
  * `age`: The age of the patient at the time of egg retrieval (units: years)
  * `dose`: The total daily dose of recombinant follicle-stimulating hormone (rFSH) administered during ovarian stimulation (IU/day)
  * `prev_failure`: Whether the patient had at least one previous failed IVF cycle (1 = yes, 0 = no)

* **pollen_data.csv**: Comes from a field experiment examining how nitrogen enrichment and plant biomass influence grassland pollinator abundance. Plots were surveyed between July 15th and 30th, 2006 when pollinator visits were counted. Each row represents a unique plot. 
  * `visits`: Number of visits from pollinators of all species recorded during the 15-minute observation window.
  * `fertilized`: Whether the plot received 10 g/m^2 nitrogen fertilizer once at the start of the season (1 = yes, 0 = no)
  * `biomass`: Aboveground plant biomass harvested at peak season (g/m^2)
  * `area`: Precise plot size surveyed, accounting for slight variations (m^2)

Load the datasets by finishing the `read_csv()` function.

```{r}
pollen_data <- read_csv()
ivf_data <- read_csv()
```

## Part 1: Logistic regression

Logistic regression tells us how the probability that X is a success in N trials, given a probability of success = p, changes as a function of some independent or explanatory variable(s). 

Using our IVF data, we are going to test the hypothesis that *aging reduces the efficacy of IVF treatment*. Our prediction is that the probability of achieving clinical pregnancy declines with age.

### Part 1a: Fit the models

We are going to jump right into model-fitting to test our hypothesis. We will use the familiar function `lm()` and a new function, `glm()`, which fits generalized linear models. `glm()` takes the same arguments as `lm()`: the model formula (`y ~ x`) and `data = `. It also has an additional argument: `family = `. This argument specifies the details of how the function should fit the model. Among others the function recognizes the following family arguments corresponding to models we've already talked about:

* Linear regression: `gaussian(link = "identity")`
* Logistic regression: `binomial(link = "logit")`
* Gamma regression: `Gamma(link = "inverse")`
* Poisson regression: `poisson(link = "log")`

Note that the family specification *should not be in quotes*. For now, we will avoid talking about the `(link = )` components of the arguments.

Maybe you recognize the term "Gaussian", which refers to our familiar normal distribution. Yes, simple linear models are just a special version of GLM! We could also fit a simple linear model using the Gaussian family and `glm()`: 

```{r, eval = FALSE}
# Simple linear model using lm()
lm(y ~ x, data = dat)

# Same model fit using glm()
glm(y ~ x, data = dat, family = gaussian(link = "identity"))
```

We know that we want to test the effect of age on pregnancy. We also know that the outcome, "pregnant", is binary (yes/no), so we're looking at logistic regression.

### **Mini Challenge 1**

Let's fit some models to test our prediction that achieving pregnancy declines with age.

1. Use the `lm()` function to fit a linear regression model testing for an effect of `age` on `pregnant`.
2. Use the `glm()` function to fit a logistic regression model testing for an effect of `age` on `pregnant`.
3. Call the models "pregnancy_lm" and "pregnancy_glm".

Use the code block below for your answer!

```{r}

```


#-------------------------------------------------

### Part 1b: Assess model fit

Now, we'll take a look at the model summaries, assessing goodness of fit and effect sizes.

* Summarize both models (looks OK)
* Plot lm() diagnostics (does not look good)
* Predict and plot the lm() model, showing how it doesn't fit the data well
* Compare models using AIC

## Part 2: Poisson regression






